An exposition on the propriety of restricted Boltzmann machines

### 2016-09-03 writing/draft.Rmd
update uninterpetability

The RBM class of models can be thought to exhibit an analogous occurence of *near-degeneracy* when random variables in the neg-potential function $Q(\cdot)$ have a mean vector $\mu(\boldsymbol \theta)$ close to the boundary of the convex hull of $\mathcal{S}$, where $t(\boldsymbol x) = \{v_1, \dots, v_V, h_1, \dots, h_H, v_1 h_1, \dots, v_V h_H \} \in \mathcal{S}$ and the mean parameterization on the model parameters, $\mu(\boldsymbol \theta)$ is
\begin{align*}
\mu(\boldsymbol \theta) &= \text{E}_{\boldsymbol \theta}\left[ t(\boldsymbol X) \right] \\
&= \sum\limits_{x \in \mathcal{X}} \left\{ t(\boldsymbol x) f_{\boldsymbol \theta} (\boldsymbol x) \right\} \\
&= \sum\limits_{x \in \mathcal{X}} \left\{ t(\boldsymbol x)\dfrac{\exp\left(\sum\limits_{i = 1}^V \sum\limits_{j=1}^H \theta_{ij} v_i h_j + \sum\limits_{i = 1}^V\theta_{v_i} v_i + \sum\limits_{j = 1}^H\theta_{h_j} h_j\right)}{\sum\limits_{\boldsymbol x \in \mathcal{X}}\exp\left(\sum\limits_{i = 1}^V \sum\limits_{j=1}^H \theta_{ij} v_i h_j + \sum\limits_{i = 1}^V\theta_{v_i} v_i + \sum\limits_{j = 1}^H\theta_{h_j} h_j\right)}\right\} \\
\end{align*}

@@

### Uninterpretability
@kaiser2007statistical defines a measure of model impropriety called *uninterpretability*, which is characterized as due to the existence of dependence, marginal mean-structure no longer being maintained in the model. This measure concerns the difference between model expectations, E$\left[\boldsymbol X | \boldsymbol \theta\right]$, and expectations given independence, E$\left[\boldsymbol X | \emptyset\right ]$. Using this concept of dependence in a model, it is possible to understand what conditions lead to uninterpretability in a model versus those that guarantee interpretable models. In this method, a model is considered without uninterpretability if its expected value is not very different from the same model with assumed independence. The quantities to compare in the RBM case are

\begin{align*}
\text{E}\left[\boldsymbol X | \boldsymbol \theta\right] &= \sum\limits_{\boldsymbol x \in \mathcal{X}} \boldsymbol x f_{\boldsymbol \theta}(\boldsymbol x) \\
&= \sum\limits_{\boldsymbol x \in \mathcal{X}} \boldsymbol x \frac{\exp\left(\sum\limits_{i = 1}^V \sum\limits_{j=1}^H \theta_{ij} v_i h_j + \sum\limits_{i = 1}^V\theta_{v_i} v_i + \sum\limits_{j = 1}^H\theta_{h_j} h_j\right)}{\sum\limits_{\boldsymbol x \in \mathcal{X}}\exp\left(\sum\limits_{i = 1}^V \sum\limits_{j=1}^H \theta_{ij} v_i h_j + \sum\limits_{i = 1}^V\theta_{v_i} v_i + \sum\limits_{j = 1}^H\theta_{h_j} h_j\right)} \\
&= \frac{\sum\limits_{\boldsymbol x \in \mathcal{X}} x_k \exp\left(\sum\limits_{i = 1}^V \sum\limits_{j=1}^H \theta_{ij} v_i h_j + \sum\limits_{i = 1}^V\theta_{v_i} v_i + \sum\limits_{j = 1}^H\theta_{h_j} h_j\right)}{\sum\limits_{\boldsymbol x \in \mathcal{X}}\exp\left(\sum\limits_{i = 1}^V \sum\limits_{j=1}^H \theta_{ij} v_i h_j + \sum\limits_{i = 1}^V\theta_{v_i} v_i + \sum\limits_{j = 1}^H\theta_{h_j} h_j\right)}  \qquad k = 1, \dots, H + V
\end{align*}
and
\begin{align*}
\text{E}\left[\boldsymbol X | \emptyset \right] &= \sum\limits_{\boldsymbol x \in \mathcal{X}} \boldsymbol x \frac{\exp\left(\sum\limits_{i = 1}^V \theta_{v_i} v_i + \sum\limits_{j = 1}^H\theta_{h_j} h_j\right)}{\sum\limits_{\boldsymbol x \in \mathcal{X}}\exp\left(\sum\limits_{i = 1}^V\theta_{v_i} v_i + \sum\limits_{j = 1}^H\theta_{h_j} h_j\right)} \\
&= \frac{\sum\limits_{\boldsymbol x \in \mathcal{X}} x_k \exp\left(\sum\limits_{i = 1}^V\theta_{v_i} v_i + \sum\limits_{j = 1}^H\theta_{h_j} h_j\right)}{\sum\limits_{\boldsymbol x \in \mathcal{X}}\exp\left(\sum\limits_{i = 1}^V\theta_{v_i} v_i + \sum\limits_{j = 1}^H\theta_{h_j} h_j\right)}  \qquad k = 1, \dots, H + V
\end{align*}

If $|\text{E}\left[\boldsymbol X | \boldsymbol \theta\right] - \text{E}\left[\boldsymbol X | \emptyset\right]$ is large then the RBM is uninterpetable.

>>>>

The RBM class of models can be thought to exhibit an analogous occurence of *near-degeneracy* when random variables in the neg-potential function $Q(\cdot)$, $t(\boldsymbol x)$, have a mean vector $\mu(\boldsymbol \theta)$ close to the boundary of the convex hull of $\mathcal{X}$, where $t(\boldsymbol x) = \{v_1, \dots, v_V, h_1, \dots, h_H, v_1 h_1, \dots, v_V h_H \} \in \mathcal{X}$ and the mean parameterization on the model parameters, $\mu(\boldsymbol \theta)$ is
\begin{align*}
\mu(\boldsymbol \theta) &= \text{E}_{\boldsymbol \theta} t(\boldsymbol X)  \\
&= \sum\limits_{x \in \mathcal{X}} \left\{ t(\boldsymbol x) f_{\boldsymbol \theta} (\boldsymbol x) \right\} \\
&= \sum\limits_{x \in \mathcal{X}} \left\{ t(\boldsymbol x)\dfrac{\exp\left(\sum\limits_{i = 1}^V \sum\limits_{j=1}^H \theta_{ij} v_i h_j + \sum\limits_{i = 1}^V\theta_{v_i} v_i + \sum\limits_{j = 1}^H\theta_{h_j} h_j\right)}{\sum\limits_{\boldsymbol x \in \mathcal{X}}\exp\left(\sum\limits_{i = 1}^V \sum\limits_{j=1}^H \theta_{ij} v_i h_j + \sum\limits_{i = 1}^V\theta_{v_i} v_i + \sum\limits_{j = 1}^H\theta_{h_j} h_j\right)}\right\} \\
\end{align*}

@@

### Uninterpretability
@kaiser2007statistical defines a measure of model impropriety he calls *uninterpretability*, which is characterized by marginal mean-structure not being maintained in the model. A measure of this is the magnitude of the difference between model expectations, E$\left[\boldsymbol X | \boldsymbol \theta\right]$, and expectations given independence, E$\left[\boldsymbol X | \boldsymbol \theta^* \right]$, where $\boldsymbol \theta^*$ is defined to equal $\boldsymbol \theta$ with all $\theta_{ij} = 0$ for $i = 1, \dots, V, j = 1, \dots, H$. Using this concept it is possible to investigate what conditions lead to uninterpretability in a model versus those that guarantee interpretable models. A model is considered without uninterpretability if its expected value is not very different from the same model with assumed independence. The quantities to compare in the RBM case are

\begin{align*}
\text{E}\left[\boldsymbol X | \boldsymbol \theta\right] &= \sum\limits_{\boldsymbol x \in \mathcal{X}} \boldsymbol x f_{\boldsymbol \theta}(\boldsymbol x) \\
&= \sum\limits_{\boldsymbol x \in \mathcal{X}} \boldsymbol x \frac{\exp\left(\sum\limits_{i = 1}^V \sum\limits_{j=1}^H \theta_{ij} v_i h_j + \sum\limits_{i = 1}^V\theta_{v_i} v_i + \sum\limits_{j = 1}^H\theta_{h_j} h_j\right)}{\sum\limits_{\boldsymbol x \in \mathcal{X}}\exp\left(\sum\limits_{i = 1}^V \sum\limits_{j=1}^H \theta_{ij} v_i h_j + \sum\limits_{i = 1}^V\theta_{v_i} v_i + \sum\limits_{j = 1}^H\theta_{h_j} h_j\right)} \\
&= \frac{\sum\limits_{\boldsymbol x \in \mathcal{X}} x_k \exp\left(\sum\limits_{i = 1}^V \sum\limits_{j=1}^H \theta_{ij} v_i h_j + \sum\limits_{i = 1}^V\theta_{v_i} v_i + \sum\limits_{j = 1}^H\theta_{h_j} h_j\right)}{\sum\limits_{\boldsymbol x \in \mathcal{X}}\exp\left(\sum\limits_{i = 1}^V \sum\limits_{j=1}^H \theta_{ij} v_i h_j + \sum\limits_{i = 1}^V\theta_{v_i} v_i + \sum\limits_{j = 1}^H\theta_{h_j} h_j\right)}  \qquad k = 1, \dots, H + V
\end{align*}
and
\begin{align*}
\text{E}\left[\boldsymbol X | \boldsymbol 
theta^*\right] &= \sum\limits_{\boldsymbol x \in \mathcal{X}} \boldsymbol x \frac{\exp\left(\sum\limits_{i = 1}^V \theta_{v_i} v_i + \sum\limits_{j = 1}^H\theta_{h_j} h_j\right)}{\sum\limits_{\boldsymbol x \in \mathcal{X}}\exp\left(\sum\limits_{i = 1}^V\theta_{v_i} v_i + \sum\limits_{j = 1}^H\theta_{h_j} h_j\right)} \\
&= \frac{\sum\limits_{\boldsymbol x \in \mathcal{X}} x_k \exp\left(\sum\limits_{i = 1}^V\theta_{v_i} v_i + \sum\limits_{j = 1}^H\theta_{h_j} h_j\right)}{\sum\limits_{\boldsymbol x \in \mathcal{X}}\exp\left(\sum\limits_{i = 1}^V\theta_{v_i} v_i + \sum\limits_{j = 1}^H\theta_{h_j} h_j\right)}  \qquad k = 1, \dots, H + V
\end{align*}

If $|\text{E}\left[\boldsymbol X | \boldsymbol \theta\right] - \text{E}\left[\boldsymbol X | \emptyset\right]$ is large then the RBM with parameter vector $\boldsymbol \theta$ $is uninterpetable.
